---
title: "Data Visualization"
output:
  html_document:
    toc: true
    code_folding: hide
    toc_float: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)
library(readxl)
library(lubridate)
library(leaflet)
library(plotly)
library(readr)
library(knitr)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	message = FALSE
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

Our goal in this project is to visualize and analyze the SARS-CoV-2 concentrations within the 14 Wastewater Resource Recovery Facilities(WRRFs) located in the New York City. However, the visualization process is not easily achieved as the NYC wastewater dataset does not contain longitudinal and latitudinal information about the facilities. In order to successfully map the distribution of WRRFs within NYC we need to find a dataset with locational information.We found them in an international dataset "HydroWASTE" which contains the geographic data for WRRFs in many countries, containing USA. 

- How to create a dataset which contains both SARS-CoV-2 data and geographic location data?
- what is the overall SARS-CoV-2 distribution inside wastewater across NYC over the years of the pandemic?
- How does the SARS-CoV-2 concentration vary among the 14 Wastewater Resource Recovery Facilities(WRRFs) in NYC over the years of the pandemic?
- How does the SARS-CoV-2 concentration distribute within each of the Wastewater Resource Recovery Facilities(WRRFs) over the years of the pandemic?

<br />

## Data Overview

<br />

**Overview of what our dataset contains:**

```{r load the data files, message=FALSE}
map <-
  read_csv("data/HydroWASTE_v10.csv")

head(map) %>% 
  knitr::kable(digits = 3) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"), font_size = 12) %>% 
  kableExtra::scroll_box(width = "100%", height = "300px")

wwtp <-
  read_csv("data/SARS-CoV-2_concentrations_measured_in_NYC_Wastewater_20231129.csv")

wwtp %>% 
  knitr::kable(digits = 3) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"), font_size = 12) %>% 
  kableExtra::scroll_box(width = "100%", height = "300px")
```
<br />

Load two datasets and decide what varibles we will working with in addition to do data clean and combine it into one dataset called nyc_wastewater.

<br />
```{r clean the mapping data file and rename plant names}
map_df <-
  map |>
  janitor::clean_names() |>
  select(waste_id, wwtp_name, lat_wwtp, lon_wwtp) |>
  mutate(wrrf_abbreviation = recode( wwtp_name,
    "New York C   Rockaway WPCP" = "RK",
    "New York C   Red Hook WPCP" = "RH",
    "New York C   Port Richmond WPCP" = "PR",
    "New York C   Oakwood Beach WPCP" = "OB",
    "New York C   26th Ward WPCP" = "26W",
    "New York C   Tallman Island WPCP" = "TI",
    "New York C   North River WPCP" = "NR",
    "New York C   Coney Island WPCP" = "CI",
    "New York C   Jamaica WPCP" = "JA",
    "New York C   Hunts Point WPCP" = "HP",
    "New York C   Owls Head WPCP" = "OH",
    "New York C   Bowery Bay WPCP" = "BB",
    "New York C   Newtown Creek WPCP" = "NC",
    "New York C   Wards Island WPCP" = "WI",
  ))

wwtp_df <-
  wwtp |>
  janitor:: clean_names()

merge_df <- inner_join(map_df, wwtp_df, by = "wrrf_abbreviation")

nyc_wastewater <-
  merge_df |>
  rename(concentration = concentration_sars_co_v_2_gene_target_n1_copies_l) |>
  drop_na(concentration) |>
  separate(sample_date, into = c("month", "day", "year"), convert = TRUE) %>% 
   mutate(
     year = as.character(year),
     month = factor(month, levels = 1:12),
     month = recode(month,
                        "1" = "January",
                        "2" = "February",
                        "3" = "March",
                        "4" = "April",
                        "5" = "May",
                        "6" = "June",
                        "7" = "July",
                        "8" = "August",
                        "9" = "September",
                        "10" = "October",
                        "11" = "November",
                        "12" = "December")) |>
  select(-waste_id, -test_date, -per_capita_sars_co_v_2_load_n1_copies_per_day_per_population, -population_served_estimated, -wwtp_name, -annotation, -wrrf_abbreviation)

nyc_wastewater %>% 
  knitr::kable(digits = 3) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"), font_size = 12) %>% 
  kableExtra::scroll_box(width = "100%", height = "300px")
```

<br />

Creating the dataset for an overall trend for Covid-19 concentration across the 14 waste water facilities within New York City from 2020 to 2023.

```{r}
overall_trend <-
  nyc_wastewater |>
  filter(technology == "RT-qPCR") |>
  select(-lat_wwtp, -lon_wwtp) |>
  group_by(year, month) |>
  summarise(avg_conc = mean(concentration)) |>
  unite("year_month",year, month, sep = "_") |>
  mutate(month_avg = ymd(paste0(year_month, "_01")))
```

<br />

## Overall Trend Over Pandemic

<br />

#### Monthly average Covid-19 concentration in wastewater in NY measured from 2020 to 2023

```{r}
overall_trend |>
  ggplot(aes(x = month_avg, y = avg_conc)) + 
  geom_point(size = 3, alpha = .7) +
  geom_smooth(size = 2, se = FALSE, alpha = .3) +
    labs(
      x = "Time", 
      y = "Monthly Average Wastewater COVID-19 Concentration (N/L)") + 
  theme(axis.line = element_line(color = "grey"), 
      panel.background = element_blank(), 
      legend.position = "none", 
      panel.grid.major = element_line(color = "light grey", linetype = "dashed"),
      plot.title = element_text(hjust = 0.5),
      axis.title.y = element_text(size = 9.5))
```

<br />

```{r}
rtqpcr <-
  nyc_wastewater |>
  filter(year %in% c(2021, 2022), technology == "RT-qPCR") |>
  select(-lat_wwtp, -lon_wwtp)

dpcr <-
  nyc_wastewater |>
  filter(year %in% c(2021, 2022), technology == "dPCR") |>
  select(-lat_wwtp, -lon_wwtp)
```


By comparing datasets from different techniques (RT-qPCR and dPCR), we found that "dPCR" had only few observations, which was insufficient for some analysis to show trends in Covid- 19 concentrations. So we decide to use only the dataset with technology "RT-qPCR". Also "RT-qPCR" data for 2020 and 2023 are incomplete, so we only used data from 2021 and 2022 for our analysis.

<br />

## NYC Wastewater Resource Recovery Facilities Trend

<br />

#### Average Covid-19 Concentration of each Wastewater Resource Recovery Facility in 2021 in New York City

```{r}
spaghetti_plot_2021 <-
  rtqpcr |>
  filter(year == 2021) |>
  group_by(month, wrrf_name) |>
  summarise(avg_conc = mean(concentration)) |>
  ggplot(aes(x = month, y = avg_conc, color = wrrf_name, group = wrrf_name)) +
  geom_line(alpha = .5) +
  geom_point(alpha = .5) +
  labs(x = "Month", 
       y = "Mean Covid-19 Concentration", 
       color = "Water Facility") + 
  theme(legend.position = "bottom",
        legend.box.background = element_rect(),
        legend.key.size = unit(0.001, 'cm'),
        axis.text.x = element_text(angle = 60, hjust = 1),
        legend.text = element_text(size = 6))

spaghetti_plot_2021
```

<br />

#### Average Covid-19 Concentration of each Wastewater Resource Recovery Facility in 2022 in New York City

```{r}
spaghetti_plot_2022 <-
  rtqpcr |>
  filter(year == 2022) |>
  group_by(month, wrrf_name) |>
  summarise(avg_conc = mean(concentration)) |>
  ggplot(aes(x = month, y = avg_conc, color = wrrf_name, group = wrrf_name)) +
  geom_line(alpha = .5) +
  geom_point(alpha = .5) +
  labs(x = "Month", 
       y = "Mean Covid-19 Concentration", 
       color = "Water Facility") + 
  theme(legend.position = "bottom",
        legend.box.background = element_rect(),
        legend.key.size = unit(0.001, 'cm'),
        axis.text.x = element_text(angle = 60, hjust = 1),
        legend.text = element_text(size = 6)) 

spaghetti_plot_2022
```

<br />

We decided to use boxplots to represent the distribution of COVID-19 concentration by wastewater facility because boxplots  give concise  visual summaries of the distributions by displaying key statistics such as the median, quartiles, and potential outliers.

<br />

#### Box Plot of Average Covid-19 Concentration by NYC Wastewater Resource Recovery Facility in 2021

```{r}
box_plot_2021 <- rtqpcr %>%
  filter(year == 2021) %>%
  group_by(month, wrrf_name) %>%
  summarise(avg_conc = mean(concentration)) %>%
  plot_ly(
    x = ~avg_conc,
    y = ~wrrf_name,
    type = "box",
    color = ~wrrf_name,
    colors = "viridis"
  ) %>%
  layout(
    xaxis = list(title = "Mean concentration (N/L)"),
    yaxis = list(title = "Area"),
    showlegend = FALSE
  )

box_plot_2021

```

<br />

Based on the boxplots, the Owls Head Wastewater facility has the largest interquartile range with some right-skewness, which means the average concentration varies the most in this location with outliers. The Owls Head wastewater facility has the highest median of >10,000 N/L of COVID-19 concentration while all other wastewater facilities have on average less than 10,000 N/L of COVID-19 concentration.

<br />

#### Box Plot of Average Covid-19 Concentration by NYC Wastewater Resource Recovery Facility in 2022

```{r}
box_plot_2022 <- rtqpcr %>%
  filter(year == 2022) %>%
  group_by(month, wrrf_name) %>%
  summarise(avg_conc = mean(concentration)) %>%
  plot_ly(
    x = ~avg_conc,
    y = ~wrrf_name,
    type = "box",
    color = ~wrrf_name,
    colors = "viridis"
  ) %>%
  layout(
    xaxis = list(title = "Mean concentration (N/L)"),
    yaxis = list(title = "Area"),
    showlegend = FALSE
  )

box_plot_2021
```

<br />

Though the exact numbers vary, the trends in 2022 are the same as in 2021.

<br />