---
title: "Regression and Analysis"
author: "Jianing Liu"
date: "2023-12-04"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)
library(readxl)
library(lubridate)
library(leaflet)
library(plotly)
library(readr)
library(knitr)
library(broom)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 16, 
  fig.height = 12,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```


```{r load the data files}
map <-
  read_csv("data/HydroWASTE_v10.csv")

head(map) %>% 
  knitr::kable(digits = 3) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"), font_size = 12) %>% 
  kableExtra::scroll_box(width = "100%", height = "300px")
  

wwtp <-
  read_csv("data/SARS-CoV-2_concentrations_measured_in_NYC_Wastewater_20231129.csv")

head(wwtp) %>% 
  knitr::kable(digits = 3) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"), font_size = 12) %>% 
  kableExtra::scroll_box(width = "100%", height = "300px")
```


```{r clean the mapping data file and rename plant names}
map_df <-
  map |>
  janitor::clean_names() |>
  select(waste_id, wwtp_name, lat_wwtp, lon_wwtp) |>
  mutate(wrrf_abbreviation = recode( wwtp_name,
    "New York C   Rockaway WPCP" = "RK",
    "New York C   Red Hook WPCP" = "RH",
    "New York C   Port Richmond WPCP" = "PR",
    "New York C   Oakwood Beach WPCP" = "OB",
    "New York C   26th Ward WPCP" = "26W",
    "New York C   Tallman Island WPCP" = "TI",
    "New York C   North River WPCP" = "NR",
    "New York C   Coney Island WPCP" = "CI",
    "New York C   Jamaica WPCP" = "JA",
    "New York C   Hunts Point WPCP" = "HP",
    "New York C   Owls Head WPCP" = "OH",
    "New York C   Bowery Bay WPCP" = "BB",
    "New York C   Newtown Creek WPCP" = "NC",
    "New York C   Wards Island WPCP" = "WI",
  ))
```

```{r clean the waste water data file}
wwtp_df <-
  wwtp |>
  janitor:: clean_names()
```

```{r merge the waste water data file and mapping data file}
merge_df <- inner_join(map_df, wwtp_df, by = "wrrf_abbreviation")
```

```{r choose the variables from the merged data file}
nyc_wastewater <-
  merge_df |>
  rename(concentration = concentration_sars_co_v_2_gene_target_n1_copies_l) |>
  drop_na(concentration) |>
  separate(sample_date, into = c("month", "day", "year"), convert = TRUE) %>% 
   mutate(
     year = as.character(year),
     month = factor(month, levels = 1:12),
     month = recode(month,
                        "1" = "January",
                        "2" = "February",
                        "3" = "March",
                        "4" = "April",
                        "5" = "May",
                        "6" = "June",
                        "7" = "July",
                        "8" = "August",
                        "9" = "September",
                        "10" = "October",
                        "11" = "November",
                        "12" = "December")) |>
  select(-waste_id, -test_date, -per_capita_sars_co_v_2_load_n1_copies_per_day_per_population, -population_served_estimated, -wwtp_name, -annotation, -wrrf_abbreviation, -lon_wwtp, -lat_wwtp)
```


Regression part
Manhattan, Brooklyn, Bronx, Queens, Staten Island is assigned 0-4, Manhattan is assigned as reference
```{r}
borough_wastewater <-
  nyc_wastewater |>
  filter(year %in% c(2021, 2022), technology == "RT-qPCR") |>
  group_by(month, year, wrrf_name) |>
  summarise(avg_conc = mean(concentration)) |>
  mutate(
    boroughs = case_match(wrrf_name,
                         "26th Ward" ~ "Brooklyn",
                         "Bowery Bay" ~ "Manhattan",
                         "Coney Island" ~ "Brooklyn",
                         "Hunts Point" ~ "Bronx",
                         "Jamaica Bay" ~ "Queens",
                         "Newtown Creek" ~ "Queens",
                         "North River" ~ "Manhattan",
                         "Oakwood Beach" ~ "Staten Island",
                         "Owls Head" ~ "Brooklyn",
                         "Port Richmond" ~ "Staten Island",
                         "Red Hook" ~ "Brooklyn",
                         "Rockaway" ~ "Queens",
                         "Tallman Island" ~ "Queens",
                         "Wards Island" ~ "Manhattan"
                         )
  ) |>
  mutate (borough_id = as.factor(case_match(
      boroughs,
      "Manhattan" ~ 0,
      "Brooklyn" ~ 1,
      "Bronx" ~ 2,
      "Queens" ~ 3,
      "Staten Island" ~ 4
    ))
    )
  
```

Creating dataset for regression analysis, assigned months to seasons and use summer as reference.
```{r}
data_reg =
  borough_wastewater %>% 
  mutate (season = case_match(month,
                        "January" ~ "Winter",
                        "February" ~ "Winter",
                        "March" ~ "Spring",
                        "April" ~ "Spring",
                        "May" ~ "Spring",
                        "June" ~ "Summer",
                        "July" ~ "Summer",
                        "August" ~ "Summer",
                        "September" ~ "Autumn",
                        "October" ~ "Autumn",
                        "November" ~ "Autumn",
                        "December" ~ "Winter"))
data = 
data_reg %>% 
  mutate (season_id = as.factor(case_match(
      season,
      "Summer" ~ 0,
      "Autumn" ~ 1,
      "Winter" ~ 2,
      "Spring" ~ 3
    ))
    )
```



```{r}
data_fit_logistic =
  glm(avg_conc ~ borough_id + relevel(season_id,ref = "0"), data = data, family = gaussian)

summary(data_fit_logistic)

data_fit_logistic |>
  broom::tidy() |> 
  knitr::kable(digits = 3)
```



```{r}

```


Running ANOVA

```{r Anova for boroughs}
model_anova <- aov(avg_conc ~ borough_id, data = borough_wastewater)
summary(model_anova)

```
```{r Anova for seasons}
model_anova1 <- aov(avg_conc ~ season_id, data = data_reg)
summary(model_anova1)
```


Regression models

Regression analysis


model_reg = glm(avg_conc ~ season_id + borough_id, data = data_reg)

model_dataset <-
  model_reg |>
  broom::tidy() |>
knitr::kable(digits = 3)



log_beta = model_reg |>
  nest(data = )
  broom::tidy() |>
  summarise(lower_CI = exp(estimate - 1.96 * std.error),
            upper_CI = exp(estimate + 1.96 * std.error)) |>
  knitr::kable(digits = 3)

logit_model = model_reg |>
  mutate(
logit_model = map(.x = data, ~glm(avg_conc ~ season_id + borough_id, data = data_reg)),
  logit_results = map(logit_model, broom::tidy))



confidence_interval=
  model_dataset %>%
  mutate(season_id = as.factor(season_id),
         borough_id = as.factor(borough_id),
         season_id = factor(season_id, levels = 0:3),
         borough_id = factor(borough_id, levels = 0:4)) %>% 
  summarise(OR = exp(estimate),
            lower_CI = exp(estimate - 1.96 * std.error),
            upper_CI = exp(estimate + 1.96 * std.error))
confidence_intterval 
  
```{r}
```


```{r}
```

