---
title: "Final Report"
output:
  html_document:
    toc: true
    code_folding: hide
    toc_float: true
---

## Motivation
<div class="well well-sm">

COVID-19 cases are constantly fluctuating and to better understand recent trends, we decided to analyze SARS-CoV-2 concentrations in wastewater data for the possibility of future public health interventions. New York City is a place with a highly dense population and any known sources of COVID-19 contamination easily affect and spread the disease rapidly. There is an abundance of early days research on the pandemic, but we were interested in the post-pandemic COVID-19 contamination, therefore we decided to look into NYC wastewater, shedding light on potential trends and hotspots.

</div>

<br />

## Related work

<div class="well well-sm">

According to the National Center for Disease and Control (CDC), wastewater surveillance can provide early warnings of COVID-19 spread to communities and to allow for public officials to take action more quickly. It is found that SARS-CoV-2 virus can be transferred to feces, which is a way to detect those who are contaminated even before they start to show symptoms. We were inspired by the New York State's ongoing/weekly updated COVID-19 Wastewater Surveillance system. For the purposes of our project, we thought that we can do something similar by focusing on NYC wastewater and analyze any statistically significant differences that you may not get from a government website.

</div>

<br />

## Initial questions

<div class="well well-sm">
1. What trends can we see from the NYC wastewater COVID-19 concentrations?
2. Is there a significant difference in NYC COVID-19 wastewater concentrations by season?
3. Is there a significant difference in COVID-19 wastewater concentrations by NYC borough?
4. Are seasons and NYC boroughs a significant predictor of COVID-19 wastewater concentrations?


With our interest in the current COVID-19 climate in NYC, we first create data visualizations to see what trends are apparent. We saw a clear trend in high COVID-19 concentrations in the winter time and lower in the winter times. However, we knew we needed to test this hypothesis to make any conclusions. For the purposes of this project, takign locations into consideration would allow us to demonstrate our data analysis and visualizations skills.

</div>

<br />

## Data 
<br />

### About the data

We used open data sources from 'NYC Open Data' and 'figshare':

* SARS-CoV-2 concentrations measured in NYC Wastewater (NYC Open Data)

* HydroWaste Mapping data (figshare)


For our purposes, we have loaded the following R packages:

* `tidyverse`

* `ggridges`

* `patchwork`

* `readxl`

* `lubridate`

* `leaflet`

* `readr`

* `knitr`

* `broom`

* `plotly`

* `kableExtra`


```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)
library(readxl)
library(lubridate)
library(leaflet)
library(plotly)
library(readr)
library(knitr)
library(broom)
library(kableExtra)
```


First, we imported the SARS-CoV-2 concentrations measured in NYC Wastewater and HydroWaste mapping data using the read_csv.

Here is a glimpse of the types of variables we have:

```{r message=FALSE,warning=FALSE}
map <-
  read_csv("data/HydroWASTE_v10.csv")

head(map) %>% 
  knitr::kable(digits = 3) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"), font_size = 12) %>% 
  kableExtra::scroll_box(width = "100%", height = "300px")
  

wwtp <-
  read_csv("data/SARS-CoV-2_concentrations_measured_in_NYC_Wastewater_20231129.csv")

head(wwtp) %>% 
  knitr::kable(digits = 3) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"), font_size = 12) %>% 
  kableExtra::scroll_box(width = "100%", height = "300px")
```
<br />

### Data Cleaning
```{r clean the mapping data file and rename plant names}
map_df <-
  map |>
  janitor::clean_names() |>
  select(waste_id, wwtp_name, lat_wwtp, lon_wwtp) |>
  mutate(wrrf_abbreviation = recode( wwtp_name,
    "New York C   Rockaway WPCP" = "RK",
    "New York C   Red Hook WPCP" = "RH",
    "New York C   Port Richmond WPCP" = "PR",
    "New York C   Oakwood Beach WPCP" = "OB",
    "New York C   26th Ward WPCP" = "26W",
    "New York C   Tallman Island WPCP" = "TI",
    "New York C   North River WPCP" = "NR",
    "New York C   Coney Island WPCP" = "CI",
    "New York C   Jamaica WPCP" = "JA",
    "New York C   Hunts Point WPCP" = "HP",
    "New York C   Owls Head WPCP" = "OH",
    "New York C   Bowery Bay WPCP" = "BB",
    "New York C   Newtown Creek WPCP" = "NC",
    "New York C   Wards Island WPCP" = "WI",
  ))
```
<br />

The dataset contains SARS-COV-2 concentration(N1/L) tested from various WRRFs in NYC from late 2020 to 2023 using RT-qPCR and dPCR techniques. We decided to place our focus on analyzing COVID-19 concentrations of the year 2021 and 2022 as these are most relevant to understand the SARS-COV-2 pandemic. As technology advances, scientists are gradually shifting their testing focus from RT-qPCR to dPCR technique. It would be interesting to compare the significances between the two testing technologies, however, dPCR do not have enough sample to be considered during the period 2020-2022. We are able to create visualizations with the 2023 observations, but we decided to leave it out of our final analysis because data was only taken from January to April of 2023. Therefore we only focus on RT-qPCR testing results collected during the pandemic (2021 to 2022). 

<br />
```{r clean the waste water data file}
wwtp_df <-
  wwtp |>
  janitor:: clean_names()

merge_df <- inner_join(map_df, wwtp_df, by = "wrrf_abbreviation")
```
<br />



<br />
```{r choose the variables from the merged data file, message=FALSE}
nyc_wastewater <-
  merge_df |>
  rename(concentration = concentration_sars_co_v_2_gene_target_n1_copies_l) |>
  drop_na(concentration) |>
  separate(sample_date, into = c("month", "day", "year"), convert = TRUE) %>% 
   mutate(
     year = as.character(year),
     month = factor(month, levels = 1:12),
     month = recode(month,
                        "1" = "January",
                        "2" = "February",
                        "3" = "March",
                        "4" = "April",
                        "5" = "May",
                        "6" = "June",
                        "7" = "July",
                        "8" = "August",
                        "9" = "September",
                        "10" = "October",
                        "11" = "November",
                        "12" = "December")) |>
  select(-waste_id, -test_date, -per_capita_sars_co_v_2_load_n1_copies_per_day_per_population, -population_served_estimated, -wwtp_name, -annotation, -wrrf_abbreviation)

head(nyc_wastewater) %>% 
  knitr::kable(digits = 3) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"), font_size = 12) %>% 
  kableExtra::scroll_box(width = "100%", height = "300px")
```
<br />

The raw mapping dataset (HydroWASTE_v10) has 58503 observations and 25 variables, including nations, WWTP(Waste Water Treatment Plant) names, locational data, abrreations, population served by individual WTTP, and many more. For our mapping purposes, we only kept 4 variables including waste_id, wttp_name, lat_wttp, lon_wttp to keep track of the WWTP specifically those located in New York City. 
<br />
The raw wastewater dataset has 4270 observations and 9 variables. For our purposes, we cleaned and extracted the test date to month, year, and kept wrrf_name, concentration (n1 copies/L), and testing technology used (RT-qPCR / dPCR).

Our final merged dataset (mapping and wastewater) for analysis is called **nyc_wastewater** with 4129 observations, and 8 variables. Those include lat_wttp, lon_wttp, month, day, year, wrrf_name, concentration, technology.

<br />

**Data dictionary:**


* `lat_wttp`: latitude of water facility location

* `lon_wttp`: longitude of water facility location

* `month`: test date (month)

* `day`: test date (day)

* `year`: test date (year)

* `wrrf_name`: Wastewater Resource Recovery Facility (waste water treatment plant) where sample was taken

* `concentration`: Concentration of the N1 target of SARS-CoV2 genetic material measured in wastewater influent.

* `technology`: COVID-19 concentration test detection method (either RT-qPCR or dPCR)


<br />

Here, we are reading the dataset for an overall trend for Covid-19 concentration across the 14 waste water facilities within New York City from 2020 to 2023.

<br />
```{r message=FALSE, warning=FALSE}
overall_trend <-
  nyc_wastewater |>
  filter(technology == "RT-qPCR") |>
  select(-lat_wwtp, -lon_wwtp) |>
  group_by(year, month) |>
  summarise(avg_conc = mean(concentration)) |>
  unite("year_month",year, month, sep = "_") |>
  mutate(month_avg = ymd(paste0(year_month, "_01")))
```

<br />
```{r subdatset, message=FALSE, warning=FALSE}
rtqpcr <-
  nyc_wastewater |>
  filter(year %in% c(2021, 2022), technology == "RT-qPCR") |>
  select(-lat_wwtp, -lon_wwtp)
# rtqpcr dataset with 2387 observations and 6 variables: month, day, year, wrrf_name, concentration, technology

dpcr <-
  nyc_wastewater |>
  filter(year %in% c(2021, 2022), technology == "dPCR") |>
  select(-lat_wwtp, -lon_wwtp)
```


<br />

## Exploratory Analysis

Let's take a look at some graphical visualizations to identify any trends.

<br />

### Overall Trend Over Pandemic

<br />

#### Monthly average Covid-19 concentration in wastewater in NY measured from 2020 to 2023

```{r message=FALSE, warning=FALSE}
overall_trend |>
  ggplot(aes(x = month_avg, y = avg_conc)) + 
  geom_point(size = 3, alpha = .7) +
  geom_smooth(size = 2, se = FALSE, alpha = .3) +
    labs(
      x = "Time", 
      y = "Monthly Average Wastewater COVID-19 Concentration (N/L)") + 
  theme(axis.line = element_line(color = "grey"), 
      panel.background = element_blank(), 
      legend.position = "none", 
      panel.grid.major = element_line(color = "light grey", linetype = "dashed"),
      plot.title = element_text(hjust = 0.5),
      axis.title.y = element_text(size = 9.5))
```

<br />

```{r message=FALSE, warning=FALSE}
rtqpcr <-
  nyc_wastewater |>
  filter(year %in% c(2021, 2022), technology == "RT-qPCR") |>
  select(-lat_wwtp, -lon_wwtp)

dpcr <-
  nyc_wastewater |>
  filter(year %in% c(2021, 2022), technology == "dPCR") |>
  select(-lat_wwtp, -lon_wwtp)
```


By comparing datasets from different techniques (RT-qPCR and dPCR), we found that "dPCR" had only few observations, which was insufficient for some analysis to show trends in Covid- 19 concentrations. So we decide to use only the dataset with technology "RT-qPCR". Also "RT-qPCR" data for 2020 and 2023 are incomplete, so we only used data from 2021 and 2022 for our analysis.

<br />

### NYC WRRFs Trend

<br />

Spaghetti plots are useful for visualizing overall trends across the 14 water facility (WRRF) groups because they display individual trajectories for each group on the same graph. We would also be able to make comparisons of patterns and variations over time. 

<br />

#### Average Covid-19 Concentration of each Wastewater Resource Recovery Facility in 2021 in New York City

```{r message=FALSE, warning=FALSE}
spaghetti_plot_2021 <-
  rtqpcr |>
  filter(year == 2021) |>
  group_by(month, wrrf_name) |>
  summarise(avg_conc = mean(concentration)) |>
  ggplot(aes(x = month, y = avg_conc, color = wrrf_name, group = wrrf_name)) +
  geom_line(alpha = .5) +
  geom_point(alpha = .5) +
  labs(x = "Month", 
       y = "Mean Covid-19 Concentration (N1/L)", 
       color = "Water Facility") + 
  theme(legend.position = "bottom",
        legend.box.background = element_rect(),
        legend.key.size = unit(0.001, 'cm'),
        axis.text.x = element_text(angle = 60, hjust = 1),
        legend.text = element_text(size = 6))

spaghetti_plot_2021
```

<br />

Based on the spaghetti plot in 2021, all water facilities follow a clear trend. The summer time months (May to July) have lowest mean wastewater COVID-19 concentrations, while the winter months (December and January) have the highest mean wastewater COVID-19 concentration. The highest mean concentration goes up to 50,000 N1/L and the lowest means go down to ~2000 N1/L.

<br />

#### Average Covid-19 Concentration of each Wastewater Resource Recovery Facility in 2022 in New York City

```{r message=FALSE, warning=FALSE}
spaghetti_plot_2022 <-
  rtqpcr |>
  filter(year == 2022) |>
  group_by(month, wrrf_name) |>
  summarise(avg_conc = mean(concentration)) |>
  ggplot(aes(x = month, y = avg_conc, color = wrrf_name, group = wrrf_name)) +
  geom_line(alpha = .5) +
  geom_point(alpha = .5) +
  labs(x = "Month", 
       y = "Mean Covid-19 Concentration (N1/L)", 
       color = "Water Facility") + 
  theme(legend.position = "bottom",
        legend.box.background = element_rect(),
        legend.key.size = unit(0.001, 'cm'),
        axis.text.x = element_text(angle = 60, hjust = 1),
        legend.text = element_text(size = 6)) 

spaghetti_plot_2022
```

<br />

Compared to 2021, 2022 water facilities have overall higher mean wastewater COVID-19 concentrations. All water facilities follow a similar  trend. In 2022, spring (March) has the lowest mean concentration of about ~5,000 N1/L across all water facilities. The winter months ( January, December, February) have the highest mean wastewater COVID-19 concentration. However, it seems that in 2022, the mean concentrations have more variability by water facility location as shown by the graph.

<br />

We decided to use boxplots to represent the distribution of COVID-19 concentration by wastewater facility because boxplots  give concise  visual summaries of the distributions by displaying key statistics such as the median, quartiles, and potential outliers.

<br />

#### Box Plot of Average Covid-19 Concentration by NYC Wastewater Resource Recovery Facility in 2021

```{r message=FALSE, warning=FALSE}
box_plot_2021 <- rtqpcr %>%
  filter(year == 2021) %>%
  group_by(month, wrrf_name) %>%
  summarise(avg_conc = mean(concentration)) %>%
  plot_ly(
    x = ~avg_conc,
    y = ~wrrf_name,
    type = "box",
    color = ~wrrf_name,
    colors = "viridis"
  ) %>%
  layout(
    xaxis = list(title = "Mean concentration (N1/L)"),
    yaxis = list(title = "Water Facility"),
    showlegend = FALSE
  )

box_plot_2021

```

<br />

Based on the boxplots, the Owls Head Wastewater facility has the largest interquartile range with some right-skewness, which means the average concentration varies the most in this location with outliers. The Owls Head wastewater facility has the highest median of >10,000 N/L of COVID-19 concentration while all other wastewater facilities have on average less than 10,000 N/L of COVID-19 concentration.

<br />

#### Box Plot of Average Covid-19 Concentration by NYC Wastewater Resource Recovery Facility in 2022

```{r message=FALSE, warning=FALSE}
box_plot_2022 <- rtqpcr %>%
  filter(year == 2022) %>%
  group_by(month, wrrf_name) %>%
  summarise(avg_conc = mean(concentration)) %>%
  plot_ly(
    x = ~avg_conc,
    y = ~wrrf_name,
    type = "box",
    color = ~wrrf_name,
    colors = "viridis"
  ) %>%
  layout(
    xaxis = list(title = "Mean concentration (N1/L)"),
    yaxis = list(title = "Water Facility"),
    showlegend = FALSE
  )

box_plot_2022
```

<br />

The 2022 box plots seem to reflect the trends we also see from our corresponding spaghetti plot. Similar to 2021, it appears that Owls Head water facility also has the most variability in addition to Jamaica Bay. There is an overall higher mean concentration across all water facilities in 2022 with the majority of the facilities having a >10,000 N1/L COVID-19 mean concentration. Bowery Bay, Jamaica Bay, Owls Head, and Tallman Island water facilities have >20,000 N1/L COVID-19 mean concentration.

<br />

### Interactive Map
```{r load the data files, message=FALSE, warning=FALSE}
map <-
  read_csv("data/HydroWASTE_v10.csv")
  

wwtp <-
  read_csv("data/SARS-CoV-2_concentrations_measured_in_NYC_Wastewater_20231129.csv")

map_df <-
  map |>
  janitor::clean_names() |>
  select(waste_id, wwtp_name, lat_wwtp, lon_wwtp) |>
  mutate(wrrf_abbreviation = recode( wwtp_name,
    "New York C   Rockaway WPCP" = "RK",
    "New York C   Red Hook WPCP" = "RH",
    "New York C   Port Richmond WPCP" = "PR",
    "New York C   Oakwood Beach WPCP" = "OB",
    "New York C   26th Ward WPCP" = "26W",
    "New York C   Tallman Island WPCP" = "TI",
    "New York C   North River WPCP" = "NR",
    "New York C   Coney Island WPCP" = "CI",
    "New York C   Jamaica WPCP" = "JA",
    "New York C   Hunts Point WPCP" = "HP",
    "New York C   Owls Head WPCP" = "OH",
    "New York C   Bowery Bay WPCP" = "BB",
    "New York C   Newtown Creek WPCP" = "NC",
    "New York C   Wards Island WPCP" = "WI",
  ))

wwtp_df <-
  wwtp |>
  janitor:: clean_names()

merge_df <- inner_join(map_df, wwtp_df, by = "wrrf_abbreviation")

nyc_wastewater <-
  merge_df |>
  rename(concentration = concentration_sars_co_v_2_gene_target_n1_copies_l) |>
  drop_na(concentration) |>
  separate(sample_date, into = c("month", "day", "year"), convert = TRUE) %>% 
   mutate(
     year = as.character(year),
     month = factor(month, levels = 1:12),
     month = recode(month,
                        "1" = "January",
                        "2" = "February",
                        "3" = "March",
                        "4" = "April",
                        "5" = "May",
                        "6" = "June",
                        "7" = "July",
                        "8" = "August",
                        "9" = "September",
                        "10" = "October",
                        "11" = "November",
                        "12" = "December")) |>
  select(-waste_id, -test_date, -per_capita_sars_co_v_2_load_n1_copies_per_day_per_population, -population_served_estimated, -wwtp_name, -annotation, -wrrf_abbreviation)

wttp_map <-
  nyc_wastewater %>% 
  filter(year %in% c(2021, 2022), technology == "RT-qPCR") %>% 
  group_by(wrrf_name) %>%
  summarise(avg_conc_area = mean(concentration))
            
map_df <- left_join(nyc_wastewater, wttp_map, by = "wrrf_name")
```


<br />

Our objective is to generate a visual representation of SARS-CoV-2 concentrations in wastewater across various Wastewater Resource Recovery Facilities (WRRFs) in New York City.

The first part of the code involves data loading and cleaning. For the WRRF information data, relevant columns are selected and prepared for further analysis. Abbreviations for WRRF names are established for convenient reference. Concentration data is also loaded and cleaned, and the two datasets are merged based on the WRRF abbreviation. In the merged dataset, date information is separated into day, month, and year columns, and the dataset is filtered to drop rows with missing concentration values. 

```{r message=FALSE, warning=FALSE}
map <-
  map_df %>% 
  leaflet() |>
  addTiles() |>
  addCircleMarkers(
             lng = ~lon_wwtp,
             lat = ~lat_wwtp,
             radius = 7,
             stroke = TRUE,
             fillOpacity = 0,
             color = "skyblue",
             popup = ~paste(sep = "<br/>",
                            "<b>", wrrf_name,
                            "</b><br>Mean Concentration:", avg_conc_area, "(N1 Copies/L)")) %>% 
  addProviderTiles(providers$CartoDB.Positron)

map
```

<br />

A map is constructed using the Leaflet package. Initially, the average concentration for each WRRF is aggregated for the years 2021 and 2022, considering only data from the "RT-qPCR" technology. The original dataset is then left-joined with this summary data based on the WRRF name.

The Leaflet map is generated using the resulting dataframe, featuring tiles from the CartoDB Positron map as a base layer. Circle markers are incorporated to denote each WRRF's geographical location, and the radius and color of the circles are adjusted for enhanced visualization. Popup information is included, showcasing the WRRF name and the mean concentration of SARS-CoV-2 (N1/L).

The map offers an vivid depiction of the distribution of SARS-CoV-2 concentrations across different WRRFs in New York City, allowing for visual distinctions among areas with higher or lower concentrations, and the popup information facilitates more detailed insights into specific WRRFs.

<br />

## Additional analysis
<br />

```{r load the analysis data files, message=FALSE, warning=FALSE}
borough_wastewater <-
  nyc_wastewater |>
  filter(year %in% c(2021, 2022), technology == "RT-qPCR") |>
  group_by(month, year, wrrf_name) |>
  summarise(avg_conc = mean(concentration)) |>
  mutate(
    boroughs = case_match(wrrf_name,
                         "26th Ward" ~ "Brooklyn",
                         "Bowery Bay" ~ "Manhattan",
                         "Coney Island" ~ "Brooklyn",
                         "Hunts Point" ~ "Bronx",
                         "Jamaica Bay" ~ "Queens",
                         "Newtown Creek" ~ "Queens",
                         "North River" ~ "Manhattan",
                         "Oakwood Beach" ~ "Staten Island",
                         "Owls Head" ~ "Brooklyn",
                         "Port Richmond" ~ "Staten Island",
                         "Red Hook" ~ "Brooklyn",
                         "Rockaway" ~ "Queens",
                         "Tallman Island" ~ "Queens",
                         "Wards Island" ~ "Manhattan"
                         )
  ) |>
  mutate (borough_id = as.factor(case_match(
      boroughs,
      "Manhattan" ~ 0,
      "Brooklyn" ~ 1,
      "Bronx" ~ 2,
      "Queens" ~ 3,
      "Staten Island" ~ 4
    ))
    )
  
data_reg =
  borough_wastewater %>% 
  mutate (season = case_match(month,
                        "January" ~ "Winter",
                        "February" ~ "Winter",
                        "March" ~ "Spring",
                        "April" ~ "Spring",
                        "May" ~ "Spring",
                        "June" ~ "Summer",
                        "July" ~ "Summer",
                        "August" ~ "Summer",
                        "September" ~ "Autumn",
                        "October" ~ "Autumn",
                        "November" ~ "Autumn",
                        "December" ~ "Winter")) %>% 
  mutate (season_id = as.factor(case_match(
      season,
      "Summer" ~ 0,
      "Autumn" ~ 1,
      "Winter" ~ 2,
      "Spring" ~ 3,
    ))
    )

```
<br />

We are interested in whether there are potential relationships: 1) between SARS-CoV-2 concentrations in wastewater and different wastewater plants by their borough location and 2) between SARS-CoV-2 concentrations in wastewater and different seasons across the year. We hope to describe patterns according to borough and season fluctuation using ANOVA and logistic regression analysis on this dataset and hopefully assist stakeholders on future policy-making processes. We tested SARS-CoV-2 concentrations in wastewater by grouping Wastewater Resource Recovery Facilities together with respect to their borough locations, including Manhattan, Brooklyn, Bronx, Queens, and Staten Island, with Manhattan as the reference. We then tested SARS-CoV-2 concentrations in wastewater by grouping  months with respect to four seasons, including summer, autumn, winter, and spring, with summer as the reference.

<br />

### Anova for Borough

<br />

#### Distribution of Average Covid-19 Wastewater Concentration By Borough

<br />

```{r boxplot data, message=FALSE, warning=FALSE}

borough_wastewater_box <-
  nyc_wastewater |>
  filter(year %in% c(2021, 2022), technology == "RT-qPCR") |>
  group_by(month, year, wrrf_name) |>
  summarise(avg_conc = mean(concentration)) |>
  mutate(
    boroughs = case_match(wrrf_name,
                         "26th Ward" ~ "Brooklyn",
                         "Bowery Bay" ~ "Manhattan",
                         "Coney Island" ~ "Brooklyn",
                         "Hunts Point" ~ "Bronx",
                         "Jamaica Bay" ~ "Queens",
                         "Newtown Creek" ~ "Queens",
                         "North River" ~ "Manhattan",
                         "Oakwood Beach" ~ "Staten Island",
                         "Owls Head" ~ "Brooklyn",
                         "Port Richmond" ~ "Staten Island",
                         "Red Hook" ~ "Brooklyn",
                         "Rockaway" ~ "Queens",
                         "Tallman Island" ~ "Queens",
                         "Wards Island" ~ "Manhattan"
                         )
  ) %>% 
  plot_ly(
    y = ~boroughs,
    x = ~avg_conc,
    type = "box",
    color = ~boroughs,
    colors = "viridis"
  ) %>%
  layout(
    yaxis = list(title = "Boroughs"),
    xaxis = list(title = "Average Covid-19 Wastewater Concentration"),
    showlegend = FALSE
  )

borough_wastewater_box

```


<br />

#### ANOVA for Boroughs

```{r Anova for boroughs, message=FALSE, warning=FALSE}
# Perform ANOVA test
model_anova_borough <- aov(avg_conc ~ borough_id, data = borough_wastewater)

tidy_anova_borough <-
  tidy(model_anova_borough) %>% 
  kbl() %>% 
  kable_material(
    c("striped", "hover")
  ) 

tidy_anova_borough
```

<br />

•	Between-group (Borough):
<br />
•	df (Degrees of Freedom): 4
<br />
•	sumsq (Sum of Squares): 202875085
<br />
•	meansq (Mean Square): 50718771
<br />
•	statistic (F-statistic): 0.4687958
<br />
•	p-value: 0.7586387
<br />
•	Within-group (Residuals):
<br />
•	df (Degrees of Freedom): 331
<br />
•	sumsq (Sum of Squares): 35810713645
<br />
•	meansq (Mean Square): 108189467
<br />

Since the p-value is greater than the significance level (0.469 > 0.05), we fail to reject the null hypothesis that different boroughs do not cause differences in SARS-COV-2 concentration within NYC wastewater during 2021 to 2022. This suggests that there is no significant between group difference in the mean SARS-COV-2 concentration among the 5 boroughs in NYC from 2021 to 2022. 

<br />

### Anova for Seasons

<br />

#### Distribution of Average Covid-19 Wastewater Concentration By Season

<br />

```{r, message=FALSE, warning=FALSE}

season_wastewater_box <-
  nyc_wastewater |>
  filter(year %in% c(2021, 2022), technology == "RT-qPCR") |>
  group_by(month, year, wrrf_name) |>
  summarise(avg_conc = mean(concentration)) |>
  mutate (season = case_match(month,
                        "January" ~ "Winter",
                        "February" ~ "Winter",
                        "March" ~ "Spring",
                        "April" ~ "Spring",
                        "May" ~ "Spring",
                        "June" ~ "Summer",
                        "July" ~ "Summer",
                        "August" ~ "Summer",
                        "September" ~ "Autumn",
                        "October" ~ "Autumn",
                        "November" ~ "Autumn",
                        "December" ~ "Winter")) %>% 
  plot_ly(
    y = ~season,
    x = ~avg_conc,
    type = "box",
    color = ~season,
    colors = "viridis"
  ) %>%
  layout(
    yaxis = list(title = "Season"),
    xaxis = list(title = "Average Covid-19 Wastewater Concentration"),
    showlegend = FALSE
  )

season_wastewater_box
```

<br />

#### Anova for Seasons

```{r Anova for seasons, message=FALSE, warning=FALSE}
# Perform ANOVA test
model_anova_season <- aov(avg_conc ~ season_id, data = data_reg)

tidy_anova_season <-
  tidy(model_anova_season) %>% 
  kbl() %>% 
  kable_material(
    c("striped", "hover")
  ) 

tidy_anova_season
```

<br />

•	Between-group (Seasons):
<br />
•	df (Degrees of Freedom): 3
<br />
•	sumsq (Sum of Squares): 8191721366
<br />
•	meansq (Mean Square): 2730573789
<br />
•	statistic (F-statistic): 32.5841
<br />
•	p-value: 1.7e-18
<br />
•	Within-group (Residuals):
<br />
•	df (Degrees of Freedom): 332
<br />
•	sumsq (Sum of Squares): 27821867364
<br />
•	meansq (Mean Square): 83800805
<br />

Since the p-value is less than the significance level (<< 0.05), we reject the null hypothesis that different seasons do not cause differences in SARS-COV-2 concentration within NYC wastewater during 2021 to 2022. This suggests that there is a significant difference in the mean SARS-COV-2 concentration among the 4 seasons from 2021 to 2022. However, the ANOVA test does not tell us which specific groups are different.

<br />

### Logistic Regression

```{r, message=FALSE, warning=FALSE}
data_fit_logistic <-
  glm(avg_conc ~ borough_id + relevel(season_id,ref = "0"), data = data_reg)

# Get summary of the logistic regression model
tidy_log <-
  tidy(data_fit_logistic) %>% 
  kbl() %>% 
  kable_material(
    c("striped", "hover")
  ) 

tidy_log
```

<br />

Then, we tried to fit regression models by using borough_id and season_id as explanatory variables and the average SARS-CoV-2 concentrations in wastewater as response variables. From the result, almost all the p-values for each categorical variable are greater than alpha = 0.05, except for season_id = 2 (indicating winter). Therefore, we can conclude that there could be an association between the average SARS-CoV-2 concentrations in wastewater and winter as an explanatory variable.

<br />

## Discussion
<br />
