---
title: "Statistical Analysis"
output:
  html_document:
    toc: true
    code_folding: hide
    toc_float: true
---

```{r setup, include=FALSE, message=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)
library(readxl)
library(lubridate)
library(leaflet)
library(plotly)
library(readr)
library(knitr)
library(broom)
library(kableExtra)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	message = FALSE
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

```{r load the data files, message=FALSE}
map <-
  read_csv("data/HydroWASTE_v10.csv")

wwtp <-
  read_csv("data/SARS-CoV-2_concentrations_measured_in_NYC_Wastewater_20231129.csv")

map_df <-
  map |>
  janitor::clean_names() |>
  select(waste_id, wwtp_name, lat_wwtp, lon_wwtp) |>
  mutate(wrrf_abbreviation = recode( wwtp_name,
    "New York C   Rockaway WPCP" = "RK",
    "New York C   Red Hook WPCP" = "RH",
    "New York C   Port Richmond WPCP" = "PR",
    "New York C   Oakwood Beach WPCP" = "OB",
    "New York C   26th Ward WPCP" = "26W",
    "New York C   Tallman Island WPCP" = "TI",
    "New York C   North River WPCP" = "NR",
    "New York C   Coney Island WPCP" = "CI",
    "New York C   Jamaica WPCP" = "JA",
    "New York C   Hunts Point WPCP" = "HP",
    "New York C   Owls Head WPCP" = "OH",
    "New York C   Bowery Bay WPCP" = "BB",
    "New York C   Newtown Creek WPCP" = "NC",
    "New York C   Wards Island WPCP" = "WI",
  ))

wwtp_df <-
  wwtp |>
  janitor:: clean_names()

merge_df <- inner_join(map_df, wwtp_df, by = "wrrf_abbreviation")

nyc_wastewater <-
  merge_df |>
  rename(concentration = concentration_sars_co_v_2_gene_target_n1_copies_l) |>
  drop_na(concentration) |>
  separate(sample_date, into = c("month", "day", "year"), convert = TRUE) %>% 
   mutate(
     year = as.character(year),
     month = factor(month, levels = 1:12),
     month = recode(month,
                        "1" = "January",
                        "2" = "February",
                        "3" = "March",
                        "4" = "April",
                        "5" = "May",
                        "6" = "June",
                        "7" = "July",
                        "8" = "August",
                        "9" = "September",
                        "10" = "October",
                        "11" = "November",
                        "12" = "December")) |>
  select(-waste_id, -test_date, -per_capita_sars_co_v_2_load_n1_copies_per_day_per_population, -population_served_estimated, -wwtp_name, -annotation, -wrrf_abbreviation, -lon_wwtp, -lat_wwtp)

borough_wastewater <-
  nyc_wastewater |>
  filter(year %in% c(2021, 2022), technology == "RT-qPCR") |>
  group_by(month, year, wrrf_name) |>
  summarise(avg_conc = mean(concentration)) |>
  mutate(
    boroughs = case_match(wrrf_name,
                         "26th Ward" ~ "Brooklyn",
                         "Bowery Bay" ~ "Manhattan",
                         "Coney Island" ~ "Brooklyn",
                         "Hunts Point" ~ "Bronx",
                         "Jamaica Bay" ~ "Queens",
                         "Newtown Creek" ~ "Queens",
                         "North River" ~ "Manhattan",
                         "Oakwood Beach" ~ "Staten Island",
                         "Owls Head" ~ "Brooklyn",
                         "Port Richmond" ~ "Staten Island",
                         "Red Hook" ~ "Brooklyn",
                         "Rockaway" ~ "Queens",
                         "Tallman Island" ~ "Queens",
                         "Wards Island" ~ "Manhattan"
                         )
  ) |>
  mutate (borough_id = as.factor(case_match(
      boroughs,
      "Manhattan" ~ 0,
      "Brooklyn" ~ 1,
      "Bronx" ~ 2,
      "Queens" ~ 3,
      "Staten Island" ~ 4
    ))
    )
  
data_reg =
  borough_wastewater %>% 
  mutate (season = case_match(month,
                        "January" ~ "Winter",
                        "February" ~ "Winter",
                        "March" ~ "Spring",
                        "April" ~ "Spring",
                        "May" ~ "Spring",
                        "June" ~ "Summer",
                        "July" ~ "Summer",
                        "August" ~ "Summer",
                        "September" ~ "Autumn",
                        "October" ~ "Autumn",
                        "November" ~ "Autumn",
                        "December" ~ "Winter")) %>% 
  mutate (season_id = as.factor(case_match(
      season,
      "Summer" ~ 0,
      "Winter" ~ 1,
      "Spring" ~ 2,
      "Autumn" ~ 3,
    ))
    )

```

<br />

We created borough_id as a categorical variable by grouping wastewater plants according to their respective five boroughs, including Manhattan, Brooklyn, Bronx, Queens, Staten Island. We set Manhattan as reference. We then created season_id as a categorical variable by grouping month according to four different seasons, including summer, autumn, winter, and spring. We set summer as the reference.

<br />

## ANOVA for Boroughs

<br />

#### Distribution of Average Covid-19 Wastewater Concentration By Borough

<br />

```{r}

borough_wastewater_box <-
  nyc_wastewater |>
  filter(year %in% c(2021, 2022), technology == "RT-qPCR") |>
  group_by(month, year, wrrf_name) |>
  summarise(avg_conc = mean(concentration)) |>
  mutate(
    boroughs = case_match(wrrf_name,
                         "26th Ward" ~ "Brooklyn",
                         "Bowery Bay" ~ "Manhattan",
                         "Coney Island" ~ "Brooklyn",
                         "Hunts Point" ~ "Bronx",
                         "Jamaica Bay" ~ "Queens",
                         "Newtown Creek" ~ "Queens",
                         "North River" ~ "Manhattan",
                         "Oakwood Beach" ~ "Staten Island",
                         "Owls Head" ~ "Brooklyn",
                         "Port Richmond" ~ "Staten Island",
                         "Red Hook" ~ "Brooklyn",
                         "Rockaway" ~ "Queens",
                         "Tallman Island" ~ "Queens",
                         "Wards Island" ~ "Manhattan"
                         )
  ) %>% 
  plot_ly(
    y = ~boroughs,
    x = ~avg_conc,
    type = "box",
    color = ~boroughs,
    colors = "viridis"
  ) %>%
  layout(
    yaxis = list(title = "Boroughs"),
    xaxis = list(title = "Average Covid-19 Wastewater Concentration"),
    showlegend = FALSE
  )

borough_wastewater_box

```


<br />

#### ANOVA for Boroughs

```{r Anova for boroughs, message=FALSE, warning=FALSE}
# Perform ANOVA test
model_anova_borough <- aov(avg_conc ~ borough_id, data = borough_wastewater)

tidy_anova_borough <-
  tidy(model_anova_borough) %>% 
  kbl() %>% 
  kable_material(
    c("striped", "hover")
  ) 

tidy_anova_borough
```

<br />

•	Between-group (Borough):
<br />
•	df (Degrees of Freedom): 4
<br />
•	sumsq (Sum of Squares): 202875085
<br />
•	meansq (Mean Square): 50718771
<br />
•	statistic (F-statistic): 0.4687958
<br />
•	p-value: 0.7586387
<br />
•	Within-group (Residuals):
<br />
•	df (Degrees of Freedom): 331
<br />
•	sumsq (Sum of Squares): 35810713645
<br />
•	meansq (Mean Square): 108189467
<br />

Since the p-value is greater than the significance level (0.469 > 0.05), we fail to reject the null hypothesis that different boroughs do not cause differences in SARS-COV-2 concentration within NYC wastewater during 2021 to 2022. This suggests that there is no significant between group difference in the mean SARS-COV-2 concentration among the 5 boroughs in NYC from 2021 to 2022. 

<br />

## ANOVA for Seasons

<br />

#### Distribution of Average Covid-19 Wastewater Concentration By Season

<br />

```{r}

season_wastewater_box <-
  nyc_wastewater |>
  filter(year %in% c(2021, 2022), technology == "RT-qPCR") |>
  group_by(month, year, wrrf_name) |>
  summarise(avg_conc = mean(concentration)) |>
  mutate (season = case_match(month,
                        "January" ~ "Winter",
                        "February" ~ "Winter",
                        "March" ~ "Spring",
                        "April" ~ "Spring",
                        "May" ~ "Spring",
                        "June" ~ "Summer",
                        "July" ~ "Summer",
                        "August" ~ "Summer",
                        "September" ~ "Autumn",
                        "October" ~ "Autumn",
                        "November" ~ "Autumn",
                        "December" ~ "Winter")) %>% 
  plot_ly(
    y = ~season,
    x = ~avg_conc,
    type = "box",
    color = ~season,
    colors = "viridis"
  ) %>%
  layout(
    yaxis = list(title = "Season"),
    xaxis = list(title = "Average Covid-19 Wastewater Concentration"),
    showlegend = FALSE
  )

season_wastewater_box
```

<br />

#### Anova for Seasons

```{r Anova for seasons, message=FALSE}
# Perform ANOVA test
model_anova_season <- aov(avg_conc ~ season_id, data = data_reg)

tidy_anova_season <-
  tidy(model_anova_season) %>% 
  kbl() %>% 
  kable_material(
    c("striped", "hover")
  ) 

tidy_anova_season
```

<br />

•	Between-group (Seasons):
<br />
•	df (Degrees of Freedom): 3
<br />
•	sumsq (Sum of Squares): 8191721366
<br />
•	meansq (Mean Square): 2730573789
<br />
•	statistic (F-statistic): 32.5841
<br />
•	p-value: ~0
<br />
•	Within-group (Residuals):
<br />
•	df (Degrees of Freedom): 332
<br />
•	sumsq (Sum of Squares): 27821867364
<br />
•	meansq (Mean Square): 83800805
<br />

Since the p-value is less than the significance level (<< 0.05), we reject the null hypothesis that different seasons do not cause differences in SARS-COV-2 concentration within NYC wastewater during 2021 to 2022. This suggests that there is a significant difference in the mean SARS-COV-2 concentration among the 4 seasons from 2021 to 2022. However, the ANOVA test does not tell us which specific groups are different.

<br />

## Logistic Regression

```{r, message=FALSE, warning=FALSE}
data_fit_logistic <-
  glm(avg_conc ~ borough_id + relevel(season_id,ref = "0"), data = data_reg)

# Get summary of the logistic regression model
tidy_log <-
  tidy(data_fit_logistic) %>% 
  kbl() %>% 
  kable_material(
    c("striped", "hover")
  ) 

tidy_log
```


<br />


In addition, we tried to fit a regression model by using borough_id and season_id as explanatory variables and the average wastewater covid-19 concentration as response variable. Almost all the p-values for each categorical variable are greater than alpha = 0.05, except for season_id = 2, indicating winter. Therefore, we can conclude that there could be a trend between the average concentration of wastewater covid-19 cases with winter season.